---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rasterbc

<!-- badges: start -->
<!-- badges: end -->

rasterbc provides access to a large collection of spatial ecological data on the province of BC during the period 2001-2018, in a common (gridded) raster format in the standard [BC Albers](https://spatialreference.org/ref/epsg/nad83-bc-albers/) projection. The goal is to make this data more accessible and easier to import into R, for the benefit of modelers interested in the forest ecology of BC.

Links to metadata associated with these layers, and code for downloading them from their original sources can be found in a sister repository, [rasterbc_src](https://github.com/deankoch/rasterbc_src). Links to the documentation for this code, and to the original sources, are listed below:

* [BC biogeoclimatic zone](https://github.com/deankoch/rasterbc_src/blob/master/src_bgcz.knit.md) ('bgcz'), from the [BC Ministry of Forests](https://catalogue.data.gov.bc.ca/dataset/f358a53b-ffde-4830-a325-a5a03ff672c3)
* [Geographical coordinates grid](https://github.com/deankoch/rasterbc_src/blob/master/src_borders.knit.md) ('borders'), from [Natural Resources Canada](https://www.nrcan.gc.ca/maps-tools-publications/maps/topographic-maps/10995)
* [Consolidated cutblocks, 2001-2018](https://github.com/deankoch/rasterbc_src/blob/master/src_cutblocks.knit.md) ('cutblocks'), from the [BC Ministry of Forests](https://catalogue.data.gov.bc.ca/dataset/harvested-areas-of-bc-consolidated-cutblocks-)
* [Digital elevation model](https://github.com/deankoch/rasterbc_src/blob/master/src_dem.knit.md) ('dem') from [Natural Resources Canada](http://ftp.geogratis.gc.ca/pub/nrcan_rncan/elevation/cdem_mnec/doc/CDEM_en.pdf)
* [Forest insect and disease survey, 2001-2018](https://github.com/deankoch/rasterbc_src/blob/master/src_fids.knit.md) ('fids'), from the [BC Ministry of Forests](https://catalogue.data.gov.bc.ca/dataset/pest-infestation-polygons)
* [Forest extent and change, 2001-2019](https://github.com/deankoch/rasterbc_src/blob/master/src_gfc.knit.md) ('gfc'), from [Hansen et al., 2013](https://www.nrcresearchpress.com/doi/full/10.1139/cjfr-2013-0401)
* [Canadian national fire database, 2001-2018](https://github.com/deankoch/rasterbc_src/blob/master/src_nfdb.knit.md) ('nfdb'), from [Natural Resources Canada](https://cwfis.cfs.nrcan.gc.ca/ha/nfdb)
* [Interpolated forest attributes, 2001, 2011](https://github.com/deankoch/rasterbc_src/blob/master/src_pine.knit.md) ('pine') from [Beaudoin et al. (2017)](https://www.nrcresearchpress.com/doi/full/10.1139/cjfr-2017-0184)

All raster data files are hosted as [a data publication](https://www.frdr-dfdr.ca/repo/handle/doi:10.20383/101.0283) (with associated [DOI](https://doi.org/10.20383/101.0283)) on [FRDR](https://www.frdr-dfdr.ca/repo/) for permanence and easy referencing. 

## News

**2021/11/16**

Updates to fix broken URLs and remove dependency on gdalUtils

**2020/09/22**

Data publication is now finalized at FRDR

**2020/09/18**

All data (including 'cutblocks') have now been uploaded to FRDR's main site. A DOI for the dataset and some updated links are forthcoming.

**2020/07/07**

Some major changes in today's commits: Function `listdata_bc` now has cleaner and more detailed information printout, with boolean return values as an optional (mostly internal) feature; `datadir_bc` arguments have been simplified; `getdata_bc` now loads/merges data by default

**2020/07/05**

All collections except 'cutblocks' are now on [FRDR's demo site](https://demo.frdr-dfdr.ca/repo/handle/doi:10.80217/demo.278?mode=full), and can be accessed via this package.

**2020/06/30**

A development version is now ready for installation from github. For now, only the 'dem' collection ('dem', 'aspect', 'slope') is accessible online from FRDR. A beta version with all layers, except possibly 'harvest' from collection 'cutblocks' (copyright permission request pending), is planned for release around July 10, 2020.

## Installation

This package is still in development, but a release on [CRAN](https://CRAN.R-project.org) is planned in the near future. For now the package may be tested by installing the `devtools` package (run `install.packages('devtools')`), and then running the following two lines:

```{r installation}
library(devtools)
install_github('deankoch/rasterbc')
```
I have tried to keep dependencies to a minimum. The package requires `sf` and `raster` for loading and merging geospatial data. If these are not already installed on your machine, the `install_github` line will ask to install them.

## Local data storage

`rasterbc` is a data-retrieval tool. Start by setting a storage directory for the raster layers:

```{r initializing}
library(rasterbc)
library(raster)
datadir_bc('H:/rasterbc_data', quiet=TRUE)
```

When `quiet=FALSE` (the default), the function will ask users to confirm that `rasterbc` should write files
to the supplied directory, and warn if this directory contains any existing files/folders. This is to caution people that if the storage directory has existing files matching those fetched by the `rasterbc` package, those data can be overwritten. Note that
files are only overwritten via calls of the form `raster::getdata_bc(..., force.dl=TRUE)` (the default is `force.dl=FALSE`).

However, to be safe you should set the data directory to a path that won't used by other applications; *eg.* a subfolder of your home directory or external storage device, with a unique folder name. In future sessions, users can set `quiet=TRUE` to skip
the interactive prompt and suppress warnings about existing data.

This path string is stored as an R option. View it using:
```{r view-datadir}
datadir_bc()
```

Depending on the geographical extent of interest and the number different layers requested, the storage demands can be high. For example, if every layer is downloaded, then around 30 GB of space is needed. Make sure you have selected a drive with enough free space for your project.

## Getting started

To demonstrate this package we'll need a polygon covering a (relatively) small geographical extent in BC. Start by loading the [bcmaps](https://github.com/bcgov/bcmaps) package and grabbing the polygons for the BC provincial boundary and the [Central Okanagan Regional District](https://www.regionaldistrict.com/)

```{r load-example, dpi=400}
library(bcmaps)

# define and load the geometry
example.name = 'Regional District of Central Okanagan'
bc.bound.sf = bc_bound()
districts.sf = regional_districts()
example.sf = districts.sf[districts.sf$ADMIN_AREA_NAME==example.name, ]

# plot against map of BC
plot(st_geometry(ntspoly_bc), main=example.name, border='red')
plot(st_geometry(bc.bound.sf), add=TRUE, col=adjustcolor('blue', alpha.f=0.2))
plot(st_geometry(example.sf), add=TRUE, col=adjustcolor('yellow', alpha.f=0.5))
text(st_coordinates(st_centroid(st_geometry(ntspoly_bc))), labels=ntspoly_bc$NTS_SNRC, cex=0.5)
```

The Okanagan polygon is shown in yellow, against a red grid that partitions the geographic extent of the province into 89 smaller regions, called *mapsheets*. This is the [NTS/SNRC](https://www.nrcan.gc.ca/maps-tools-publications/maps/topographic-maps/10995)
grid used by Natural Resources Canada for their topographic maps, with each mapsheet identied by a
[unique number-letter code](https://www.nrcan.gc.ca/earth-sciences/geography/topographic-information/maps/9765).
`rasterbc` uses this grid to package data into blocks for distribution. It is lazy-loaded as the `sf` object `ntspoly_bc`: 

```{r ntspoly_bc}
print(ntspoly_bc)
```

## A basic example

Let's download Canada's 1:250,000 digital elevation model ([CDEM](http://ftp.geogratis.gc.ca/pub/nrcan_rncan/elevation/cdem_mnec/doc/CDEM_product_specs.pdf))
layers corresponding to the yellow polygon. For the full BC extent, these rasters would occupy around 1.2GB of space. But we only want the smaller extent corresponding to the polygon. There are three blocks (totalling about 20 MB) which overlap with our region of interest

```{r finblocks-example}
findblocks_bc(example.sf)
```

fetch them using the command:

```{r get-example, eval=FALSE}
getdata_bc(geo=example.sf, collection='dem', varname='dem', load.mosaic=FALSE)
```

You should see progress bars for a series of three downloads, and once finished, the paths of the downloaded files are printed to the console. Note that if a block has been downloaded already (*eg.* by a `getdata_bc` call with a different `geo` argument), the existing copy will be detected, and the download skipped. *eg.* repeat the call...

```{r get-example-again, R.options = list(width = 15)}
getdata_bc(geo=example.sf, collection='dem', varname='dem', load.mosaic=FALSE)
```

... and nothing happens, because the data are there already. Verify by loading one of the files as `RasterLayer`:

```{r view-example, R.options = list(width = 15), dpi=400}
example.raster = raster('H:/rasterbc_data/dem/blocks/dem_092H.tif')
print(example.raster)
plot(example.raster, main='elevation (metres)')
```



### Loading/merging blocks

To display the elevation data for the entire district, we need to combine the three blocks downloaded earlier. This can be done using `getdata_bc` with `load.mosaic=TRUE` (the default setting), which loads all required blocks, merges them into a single (mosaic) layer, crops and masks as needed, and then loads into memory the returned `RasterLayer` object:
```{r opendata_bc, R.options = list(width = 10), dpi=400}
example.tif = getdata_bc(example.sf, collection='dem', varname='dem')
print(example.tif)
plot(example.tif, main=paste(example.name, ': elevation'))
plot(st_geometry(example.sf), add=TRUE)
plot(st_geometry(ntspoly_bc), add=TRUE, border='red')
```

Note that any simple features object of class `sf` or `sfc` could be used for the argument `geo`, provided its geometry intersects with the provincial boundary of BC. However, the intended usage is for the user to delineate their region of interest as a `(MULTI)POLYGON` object (here, `example.sf` is a `MULTIPOLYGON`). Note that geometries of other classes (such `SpatialPolygons`, as defined by `sp`; or data frames containing coordinates of vertices) can often be coerced to `sf` using a command like `sf::st_as_sf(other_geometry_class_object)`.

Alternatively, users can directly download individual blocks by specifying their their NTS/SNRC codes, *eg.* here is the
"slope" variable (from the "dem" collection), specified using the codes: 
```{r get-example-blocks, R.options = list(width = 10), dpi=400}
example.blockcodes = findblocks_bc(example.sf)
example.tif = getdata_bc(example.blockcodes, collection='dem', varname='slope')
plot(example.tif, main=paste('NTS/SNRC mapsheets ', paste(example.blockcodes, collapse=', '), ': slope'))
plot(st_geometry(ntspoly_bc), add=TRUE, border='red')
text(st_coordinates(st_centroid(st_geometry(ntspoly_bc))), labels=ntspoly_bc$NTS_SNRC, cex=0.5)
```


### File management

If you forget which files have been downloaded, you can either check the directory `data.dir` using your file browser (subfolder '/dem/blocks', in this case), or use `listfiles_bc` to get a logical vector indicating which files are curerntly found in your local storage directory:

```{r listdata_bc-first, R.options = list(width = 10)}
is.downloaded = listdata_bc(collection='dem', varname='dem', simple=TRUE)
str(is.downloaded)
sum(is.downloaded)
length(is.downloaded)
```

This shows that of the 89 blocks for the variable name 'dem' (in the collection 'dem'), we have downloaded three so far. Notice the return value of `listdata_bc` is a *named* vector, with names indicating the destination filenames and paths. This shows where they will be written by `getdata_bc`. All filenames are either of the form 'varname_mapsheet.tif' (as in this example) or else varname_year_mapsheet.tif (for time-series data).

By default, the `listdata_bc` function prints a list of all available layers. *eg.* in the 'dem' collection we also have 'aspect' and 'slope':

```{r listdata_bc-second, R.options = list(width = 125)}
listdata_bc(collection='dem', verbose=2)
```

Notice the 'slope' blocks that were downloaded manually using NTS/SNRC codes. We merged these blocks earlier in the `opendata_bc` function call that created `example.tif`. Currently, this layer resides 'in memory' and can be accessed via the R object `example.tif`. To save a copy, one can use the `raster::writeRaster` function. *eg.* 

```{r save-example, R.options = list(width = 15)}
slope.path = file.path(getOption('rasterbc.data.dir'), 'dem', 'example_slope.tif')
writeRaster(example.tif, slope.path, overwrite=TRUE)
```

`getdata_bc` writes all of its data inside a 'blocks' subdirectory (in this case '/dem/blocks'), and the subfolder of the data directory corresponding to the collection (in this case '/dem') is, by default, left empty. So it is a good place to store and organize such derivative files, where they can be loaded more quickly (in future), *eg.*

```{r load-saved-file, R.options = list(width = 15)}
raster(slope.path)
```

If you're finished with `rasterbc` and want to remove all of the stored data, or if you simply want to free up space, the entire data directory or any of its contents can be deleted using your file browser. This will not break the `rasterbc` installation. However, all downloaded data will be erased and you will need to run `datadir_bc` again before using the other package functions. 

### Integer codes

Note that the `bgcz` collection data are factors, which are then encoded in the geotiff files as integer codes. `opendata_bc` returns these factor names in a raster attribute table for the `RasterLayer` object (column "code"). The complete lookup
tables are also stored in the lazy loaded list object `metadata_bc`.

```{r load-lookup, R.options = list(width = 15)}
lookup.list = rasterbc::metadata_bc$bgcz$metadata$coding
print(lookup.list$zone)
```
For example we have zone, 1 = Boreal Altai Fescue Alpine (BAFA), 2 = Bunchgrass, etc. See the documentation for the
[bgcz source script](https://github.com/deankoch/rasterbc_src/blob/master/src_bgcz.knit.md) for links to a complete description of all codes.

The code below plots this data for the example region, replacing in the legend the integer levels of the
raster with the properly matched zone codes:

```{r bcgz plot}

# open the biogeoclimatic zone raster
bgcz.raster = opendata_bc(geo=example.sf, collection='bgcz', varname='zone')

# a levels table (dataframe) can be extracted with `base::levels`
bgcz.levels = levels(bgcz.raster)[[1]]

# set up a colour palette and plot with legend defined manually
bgcz.levels$color = rainbow(nrow(bgcz.levels))
plot(bgcz.raster, legend=FALSE, col=bgcz.levels$color)
legend('bottomright', legend=bgcz.levels$code, fill=bgcz.levels$color)

```

<!-- README.md is generated from README.Rmd. Please edit that file -->
<!-- rmarkdown::render('README.Rmd') -->

