---
title: "Integration with other data sources"
author: "Dean Koch"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integration with other data sources}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette, we'll see at how other data sources can be used in R alongside the layers from `rasterbc`, looking at how they can be related (*eg.* polygons from `bcmaps` that match with `rasterbc`) and how they might be different (*eg.* weather data from `daymetr`, which is projected differently). Start by loading the necessary packages:

```{r setup}
library(rasterbc)
library(bcmaps)
library(daymetr)
```

Say we're interested in the Well's Gray provincial park. First we need coordinates to delineate this area. You could find these using [openstreetmap](https://www.openstreetmap.org/) (right-click > "Show address"), or google maps (right-click > "What's here?"). 

I get the following latitude, longitude coordinates:
(top left corner) 52.8708,-120.6766
(bottom right corner) 51.7525, -119.3939
